---
title: "Programming on data.table"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming on data.table}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r init, include = FALSE}
require(data.table)
knitr::opts_chunk$set(
  comment = "#",
    error = FALSE,
     tidy = FALSE,
    cache = FALSE,
 collapse = TRUE
)
```


## Introduction

`data.table` from the very first releases enabled interface of `subset` and `with` (or `within`) functions to be used inside `[.data.frame` by defining `[.data.table` method. `subset` and `with` are base R functions that are useful to reduce repetition in code, to make it more easily readable, and usually reduce the number of total characters user has to type. Those functions are possible in R because of its quite unique feature called *lazy evaluation* that allows for a function to catch its arguments before they are evaluated, and for example to evaluate them in a scope different than scope in which they were called. Let's recap interface of `subset` function.

```{r opt_max_print_10, include = FALSE}
options(max.print = 10L) # 2 rows
```

```{r subset}
subset(iris, Species == "setosa")
```

It takes the second argument and evaluates it inside the scope of the first argument, a data.frame. It removes variable repetition, makes code more readable, making it less prone to errors.

## Problem description

Problem of this kind of interface is that we cannot easily parameterize the code that uses it. It is because expressions passed to those functions are substituted before being evaluated. Those substituted expressions are different when they are passed by user interactively from those that passed as arguments of another function.

### Example

```{r subset_error, error=TRUE}
my_subset = function(data, col, val) {
  subset(data, col == val)
}
my_subset(iris, Species, "setosa")
```

### Approaching the problem

There are multiple ways to work around the problem.

#### Avoid *lazy evaluation*

The easiest workaround is to avoid *lazy evaluation* in the first place, and fall back to less intuitive, more error-prone machnisms that uses `df[["variable"]]`, etc.

```{r subset_nolazy}
my_subset = function(data, col, val) {
  data[data[[col]] == val, ]
}
my_subset(iris, col = "Species", val = "setosa")
```

We basically compute logical vector (of length of nrow of our dataset) first, then logical vector is supplied to `i` argument of `[.data.frame` to perform ordinary subsetting by logical vector. It works well for this simple example, but it lacks flexibility, incorporates variable repetition, and requires user to change the interface to pass column name as character rather than unquoted symbol. The more complex is the expression we need to parameterize the less practical is this approach.

#### Use of `eval parse`

This method is usually preferred by newcomers to R language. Conceptually it is the most straightforward way. It is to produce required statement by using string concatenation, then parse it and evaluate.

```{r subset_parse}
my_subset = function(data, col, val) {
  data = deparse(substitute(data))
  col = deparse(substitute(col))
  val = paste0("'", val, "'")
  text = paste0("subset(", data, ", ", col, " == ", val, ")")
  eval(parse(text = text)[[1L]])
}
my_subset(iris, Species, "setosa")
```

We have to use `deparse(substitute(.))` to catch the actual names of objects passed to function, so we can construct the `subset` function call using those original names. Although It gives unlimited flexibility with a relatively low complexity, **use of `eval(parse(.))` should be avoided**. The main reasons are:

- lack of syntax validation
- [vulnerability to code injection](https://github.com/Rdatatable/data.table/issues/2655#issuecomment-376781159)
- better alternatives

Martin Machler, R Project Core Developer, [once said](https://stackoverflow.com/a/40164111/2490497):

> Sorry but I don't understand why too many people even think a string was something that could be evaluated. You must change your mindset, really. Forget all connections between strings on one side and expressions, calls, evaluation on the other side.  
The (possibly) only connection is via `parse(text = ....)` and all good R programmers should know that this is rarely an efficient or safe means to construct expressions (or calls). Rather learn more about `substitute()`, `quote()`, and possibly the power of using `do.call(substitute, ......)`.

#### Use *computing on the language*

The aforementioned functions, along with some others (including `as.call`, `as.name`/`as.symbol`, `bquote`, `eval`), can be categorized as functions to *compute on the language*, as they operate on language objects (`call`, `name`/`symbol`).

```{r subset_substitute}
my_subset = function(data, col, val) {
  eval(substitute(subset(data, col == val)))
}
my_subset(iris, Species, "setosa")
```

We used base R `substitute` function to transform call `subset(data, col == val)` into `subset(iris, Species == "setosa")` by substituting `data`, `col` and `val` with their original names (or values) from their parent environment. We can see this solution is superior to former ones. Note that we operate on the language objects layer not touching string manipulation routines, thus we refer to that process as *computing on the language*. There is a dedicated chapter on *Computing on the language* in [R language manual](https://cloud.r-project.org/doc/manuals/r-release/R-lang.html). Although it is not necessary for *programming on data.table*, we encourage readers to read this chapter, for the sake of better understanding of the R's powerful and pretty unique feature.

#### Use third party packages

There are third party packages that can achieve what base R computing on the language routines do. To name a few `pryr`, `lazyeval` or `rlang`. We will not discuss them here as they are not solving any problem that could not be addressed by base R routines.

## Programming on data.table

Now, once we established the proper way to parameterize code that uses *lazy evaluation*, we can move on to the main discussion of this vignette, *programming on data.table*.

Starting from version 1.14.1 data.table provides a robust mechanism for parameterizing expressions passed to `i`, `j` and `by` (or `keyby`) arguments of `[.data.table`. It is built upon base R `substitute` function, and mimics its interface. For that purpose `substitute2` has been added as a more robust, and more user-friendly, version of base R `substitute`. For a complete list of differences between `base::substitute` and `data.table::substitute2` please read [`substitute2` manual](https://rdatatable.gitlab.io/data.table/library/data.table/html/substitute2.html).

### Substitute variables and names

Let's assume we want to have a general function that applies a function to find the sum of two arguments that have had another function applied over them. Code of the particular example function will make this example clearer. Below we have a function to compute length of hypotenuse in a right-angled triangle, knowing length of its legs.

${\displaystyle c = \sqrt{a^2 + b^2}}$

```{r hypotenuse}
square = function(x) x^2
quote(
  sqrt(square(a) + square(b))
)
```

The goal is to make every name in the above function call passable as a *parameter*.

```{r hypotenuse_substitute2}
substitute2(
  outer(inner(var1) + inner(var2)),
  env = list(
    outer = "sqrt",
    inner = "square",
    var1 = "a",
    var2 = "b"
  )
)
```

We can see that both functions names, as well as names of the variables passed to the functions have been *parameterized*. We used `substitute2` for convenience. In this simple case, base R `substitute` could be used as well, although it requires extra `lapply(env, as.name)`.

Now, to use substitution inside `[.data.table` we don't need to call `substitute2` function. It is being used internally, all we have to do is to provide `env` argument, the same way as we provide it to `substitute2` function. Substitution is applied to `i`, `j` and `by` (or `keyby`) arguments of `[.data.table` method. Note that `verbose = TRUE` argument can be used to print expressions after substitution is applied.

```{r opt_max_print_8, include = FALSE}
options(max.print = 8L) # 2 rows
```

```{r hypotenuse_datatable}
DT = as.data.table(iris)

DT[, outer(inner(var1) + inner(var2)),
   env = list(
     outer = "sqrt",
     inner = "square",
     var1 = "Sepal.Length",
     var2 = "Sepal.Width"
  )]

# return as data.table, substitute call argument name
DT[, .(Species, var1, var2, out = outer(inner(var1) + inner(var2))),
   env = list(
     outer = "sqrt",
     inner = "square",
     var1 = "Sepal.Length",
     var2 = "Sepal.Width",
     out = "Sepal.Hypotenuse"
  )]
```

In the last call we added another parameter `out = "Sepal.Hypotenuse"` that conveys the name of output column. Unlike base R, `substitute2` will handle substitution of call arguments names as well.

Substitution works on `i` and `by` (or `keyby`) as well.

```{r hypotenuse_datatable_i_j_by}
DT[filter_col %in% filter_val,
   .(var1, var2, out = outer(inner(var1) + inner(var2))),
   by = by_col,
   env = list(
     outer = "sqrt",
     inner = "square",
     var1 = "Sepal.Length",
     var2 = "Sepal.Width",
     out = "Sepal.Hypotenuse",
     filter_col = "Species",
     filter_val = I(c("versicolor", "virginica")),
     by_col =  "Species"
  )]
```

### Substitute variables and character values

In the above example we have seen the convenience feature of `substitute2` to automatically convert characters to names/symbols. The obvious question that arises, is:  what if we actually want to substitute the parameter with a character value, so as to have a base R `substitute` behaviour? We provide a mechanism to escape this automatic conversion by wrapping elements into base R `I()` call. The `I` function marks an object as *AsIs*, preventing conversion from taking place; read `?AsIs` for more details. If base R behaviour is desired for a whole `env` argument, then best to wrap `env` argument into `I()`, otherwise each list element can be wrapped into `I()` individually, see both use cases below.

```{r rank}
substitute(    # base R
  rank(input, ties.method = ties),
  env = list(input = as.name("Sepal.Width"), ties = "first")
)

substitute2(   # like base R, env AsIs class
  rank(input, ties.method = ties),
  env = I(list(input = as.name("Sepal.Width"), ties = "first"))
)

substitute2(   # only particular elements of env are AsIs class
  rank(input, ties.method = ties),
  env = list(input = "Sepal.Width", ties = I("first"))
)
```

Note that conversion works recursively on each list element, including of course the escape mechanism.

```{r substitute2_recursive}
substitute2(   # all are symbols
  f(v1, v2),
  list(v1 = "a", v2 = list("b", list("c", "d")))
)
substitute2(   # 'a' and 'd' should stay as character
  f(v1, v2),
  list(v1 = I("a"), v2 = list("b", list("c", I("d"))))
)
```

### Substitute list of arbitrary length

Example discussed above presents neat and powerful way to make your code more dynamic. Although there are many other, much more complex cases that developer might have to deal with. One of the common problems is to handle arbitrary length list of arguments.

An obvious use case could be to mimic `.SD` functionality by injecting a `list` call into `j` argument.

```{r opt_max_print_4, include = FALSE}
options(max.print = 4L) # 2 rows
```

```{r splice_sd}
cols = c("Sepal.Length", "Sepal.Width")
DT[, .SD, .SDcols = cols]
```

We want the `cols` parameter above to be *spliced* into a `list` call making `j` argument look like below:

```{r splice_tobe}
DT[, list(Sepal.Length, Sepal.Width)]
```

*Splice* is an operation where list of objects has to be inlined into expression as a sequence of arguments to call.
In base R splice `cols` into a `list` call can be achieved using `as.call(c(quote(list), cols))`. Additionally starting from R 4.0.0, there is new interface for such operation in `bquote` function.
In data.table we make it easier, by automatically _enlist_-ing list of objects into list call to those objects. It means that any `list` object inside `env` list argument will be turned into list `call`. Making the API for that as simple as presented below.

```{r splice_datatable}
# this works
DT[, j,
   env = list(j = as.list(cols)),
   verbose = TRUE]

# this will not work
#DT[, list(cols),
#   env = list(cols = cols)]
```

It is important to provide a call to list, rather than a list, inside the `env` list argument. It is exactly what is happening in the above example, let's explain _enlist_ list into list call more in details.

```{r splice_enlist}
DT[, j,  # data.table automatically enlist nested lists into list calls
   env = list(j = as.list(cols)),
   verbose = TRUE]

DT[, j,  # turning above 'j' list into a list call
   env = list(j = quote(list(Sepal.Length, Sepal.Width))),
   verbose = TRUE]

DT[, j,  # the same as above but accepts character vector
   env = list(j = as.call(c(quote(list), lapply(cols, as.name)))),
   verbose = TRUE]
```

Now let's try to pass a list of symbols, rather than list call to those symbols. We will use `I()` to escape automatic _enlist_-ing, it will also turn off character to symbol conversion, so we also have to use `as.name`.

```{r splice_not, error=TRUE}
DT[, j,  # list of symbols
   env = I(list(j = lapply(cols, as.name))),
   verbose = TRUE]

DT[, j,  # again the proper way, enlist list to list call automatically
   env = list(j = as.list(cols)),
   verbose = TRUE]
```

While both the expressions vare visually similar, they are not identical:

```{r splice_substitute2_not}
str(substitute2(j, env = I(list(j = lapply(cols, as.name)))))

str(substitute2(j, env = list(j = as.list(cols))))
```

For more detailed explanation, please see examples in [`substitute2` manual](https://rdatatable.gitlab.io/data.table/library/data.table/html/substitute2.html).

### Substitution of a complex query

Let's take as an example a more complex function that calculates root mean square.

${\displaystyle x_{\text{RMS}}={\sqrt{{\frac{1}{n}}\left(x_{1}^{2}+x_{2}^{2}+\cdots +x_{n}^{2}\right)}}}$

It takes arbitrary number of variables on input, but now we cannot just *splice* list of arguments into a list call because each of those arguments has to be wrapped into `square` call. In this case we have to *splice* by hand rather than relying on data.table automatic _enlist_.

First we have to construct calls to `square` function for each of the variables (see `inner_calls`). Then we have to reduce list of calls into a single call, having nested sequence of `+` calls (see `add_calls`). Lastly we have to substitute constructed call into surrounding expression (see `rms`).

```{r opt_max_print_12, include = FALSE}
options(max.print = 12L) # 2 rows
```

```{r complex}
outer = "sqrt"
inner = "square"
vars = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")

syms = lapply(vars, as.name)
to_inner_call = function(var, fun) call(fun, var)
inner_calls = lapply(syms, to_inner_call, inner)
print(inner_calls)

to_add_call = function(x, y) call("+", x, y)
add_calls = Reduce(to_add_call, inner_calls)
print(add_calls)

rms = substitute2(
  expr = outer((add_calls) / len),
  env = list(
    outer = outer,
    add_calls = add_calls,
    len = length(vars)
  )
)
print(rms)

DT[, j, env = list(j = rms)]

# same but skipping last substitute2 call and using add_calls directly
DT[, outer((add_calls) / len),
   env = list(
     outer = outer,
     add_calls = add_calls,
     len = length(vars)
  )]

# return as data.table
j = substitute2(j, list(j = as.list(setNames(nm = c(vars, "Species", "rms")))))
j[["rms"]] = rms
print(j)
DT[, j, env = list(j = j)]

# alternatively
j = as.call(c(
  quote(list),
  lapply(setNames(nm = vars), as.name),
  list(Species = as.name("Species")),
  list(rms = rms)
))
print(j)
DT[, j, env = list(j = j)]
```

## Retired interfaces

In `[.data.table` it is also possible to use other interfaces for variable substitution, or passing quoted expressions. Those are `get` and `mget` for inline injection of variables by providing their names as character, and `eval` that tells `[.data.table` that expression we passed into an argument is a quoted expression, so should be handled differently. Those interfaces should be considered as retired and we recommended to use new `env` argument instead.

### `get`

```{r opt_max_print_4b, include = FALSE}
options(max.print = 4L) # 2 rows
```

```{r old_get}
v1 = "Petal.Width"
v2 = "Sepal.Width"

DT[, .(total = sum(get(v1), get(v2)))]

DT[, .(total = sum(v1, v2)),
   env = list(v1 = v1, v2 = v2)]
```

### `mget`

```{r old_mget}
v = c("Petal.Width", "Sepal.Width")

DT[, lapply(mget(v), mean)]

DT[, lapply(v, mean),
   env = list(v = as.list(v))]

DT[, lapply(v, mean),
   env = list(v = as.list(setNames(nm = v)))]
```

### `eval`

Instead of using `eval` function we can provide quoted expression into the element of `env` argument, no extra `eval` call is needed then.

```{r old_eval}
cl = quote(
  .(Petal.Width = mean(Petal.Width), Sepal.Width = mean(Sepal.Width))
)

DT[, eval(cl)]

DT[, cl, env = list(cl = cl)]
```
