---
title: "Joins in data.table"
date: "`r Sys.Date()`"
output:
  markdown::html_format
vignette: >
  %\VignetteIndexEntry{Joins in data.table}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE, message = FALSE}
require(data.table)
knitr::opts_chunk$set(
  comment = "#",
    error = FALSE,
     tidy = FALSE,
    cache = FALSE,
 collapse = TRUE
)
```

```{r, echo=FALSE, file='_translation_links.R'}
```
`r .write.translation.links("Translations of this document are available in: %s")`

In this vignette you will learn how to perform any join operation using resources available in the `data.table` syntax.

It assumes familiarity with the `data.table` syntax. If that is not the case, please read the following vignettes:

- [`vignette("datatable-intro", package="data.table")`](datatable-intro.html)
- [`vignette("datatable-reference-semantics", package="data.table")`](datatable-reference-semantics.html)
- [`vignette("datatable-keys-fast-subset", package="data.table")`](datatable-keys-fast-subset.html)

***

## 1. Defining example data

To illustrate how to use the method available with real life examples, let's simulate a **normalized database** from a little supermarket by defining the following tables in a database:

1. `Products`, a table with rows giving characteristics of various products. To show how the framework deals with ***missing values***, one `id` is `NA`.

```{r, define_products}
Products = rowwiseDT(
  id=,        name=, price=,   unit=, type=,
   1L,     "banana",   0.63,  "unit", "natural",
   2L,    "carrots",   0.89,    "lb", "natural",
   3L,    "popcorn",   2.99,  "unit", "processed",
   4L,       "soda",   1.49, "ounce", "processed",
   NA, "toothpaste",   2.99,  "unit", "processed"
)
```

2. `NewTax`, a table with rows defining some taxes associated with processed products based on their units.

```{r define_new_tax}
NewTax = data.table(
  unit = c("unit", "ounce"),
  type = "processed",
  tax_prop = c(0.65, 0.20)
)

NewTax
```


3. `ProductReceived`, a table with rows simulating weekly incoming inventory.

```{r define_product_received}
set.seed(2156)

# NB: Jan 8, 2024 is a Monday.
receipt_dates = seq(from=as.IDate("2024-01-08"), length.out=10L, by="week")

ProductReceived = data.table(
  id=1:10, # unique identifier for an supply transaction
  date=receipt_dates,
  product_id=sample(c(NA, 1:3, 6L), size=10L, replace=TRUE), # NB: product '6' is not recorded in Products above.
  count=sample(c(50L, 100L, 150L), size=10L, replace=TRUE)
)

ProductReceived
```

4. `ProductSales`, a table with rows simulating customer transactions.

```{r define_product_sales}
set.seed(5415)

# Monday-Friday (4 days later) for each of the weeks present in ProductReceived
possible_weekdays <- as.IDate(sapply(receipt_dates, `+`, 0:4))

ProductSales = data.table(
  id = 1:10,
  date = sort(sample(possible_weekdays, 10L)),
  product_id = sample(c(1:3, 7L), size = 10L, replace = TRUE), # NB: product '7' is in neither Products nor ProductReceived.
  count = sample(c(50L, 100L, 150L), size = 10L, replace = TRUE)
)


ProductSales
```

## 2. `data.table` joining syntax 

Before taking advantage of the `data.table` syntax to perform join operations we need to know which arguments can help us to perform successful joins.

The next diagram shows a description for each basic argument. In the following sections we will show how to use each of them and add more complexity little by little.

```
x[i, on, nomatch]
| |  |   |
| |  |   \__ If NULL only returns rows linked in x and i tables
| |  \____ a character vector or list defining match logic
| \_____ primary data.table, list or data.frame
\____ secondary data.table
```

> Please keep in mind that the standard argument order in `data.table` is `dt[i, j, by]`. For join operations, it is recommended to pass the `on` and `nomatch` arguments by name to avoid using `j` and `by` when they are not needed.

## 3. Equi joins

This the most common and simple case as we can find common elements between tables to combine.

The relationship between tables can be:

- **One to one**: When each matching value is unique on each table.
- **One to many**: When some matching values are repeated in one of the table both unique in the other one.
- **Many to many**: When the matching values are repeated several times on each table.

In most of the following examples we will perform *one to many* matches, but we are also going to take the time to explain the resources available to perform *many to many* matches.


### 3.1. Right join

Use this method if you need to combine columns from 2 tables based on one or more references but ***keeping all rows present in the table located on the right (in the the square brackets)***.

In our supermarket context, we can perform a right join to see more details about the products received as this is relation *one to many* by passing a vector to the `on` argument.

```{r}
Products[ProductReceived,
         on = c(id = "product_id")]
```

As many things have changed, let's explain the new characteristics in the following groups:

- **Column level**
  - The *first group* of columns in the new `data.table` comes from the `x` table.
  - The *second group* of columns in the new `data.table` comes from the `i` table.
  - If the join operation presents a present any **name conflict** (both table have same column name) the ***prefix*** `i.` is added to column names from the **right-hand table** (table on `i` position).
  
- **Row level**
  - The missing `product_id` present on the `ProductReceived` table in row 1 was successfully matched with missing `id` of the `Products` table, so `NA` ***values are treated as any other value***.
  - All rows from in the `i` table were kept including:
    - Not matching rows like the one with `product_id = 6`.
    - Rows that repeat the same `product_id` several times.
    
#### 3.1.1. Joining by a list argument

If you are following the vignette, you might have found out that we used a vector to define the relations between tables in the `on` argument, that is really useful if you are **creating your own functions**, but another alternative is to use a **list** to define the columns to match.

To use this capacity, we have 2 equivalent alternatives:

- Wrapping the related columns in the base R `list` function.

```{r, eval=FALSE}
Products[ProductReceived,
         on = list(id = product_id)]
```

- Wrapping the related columns in the `data.table` `list` alias `.`.

```{r, eval=FALSE}
Products[ProductReceived,
         on = .(id = product_id)]
```

#### 3.1.2. Alternatives to define the `on` argument

In all the prior example we have pass the column names we want to match to the `on` argument but `data.table` also have alternatives to that syntax.

- **Natural join**: Selects the columns to perform the match based on common column names. To illustrate this method, let's change the column of `Products` table from `id` to `product_id` and use the keyword `.NATURAL`.

```{r}
ProductsChangedName = setnames(copy(Products), "id", "product_id")
ProductsChangedName

ProductsChangedName[ProductReceived, on = .NATURAL]
```

- **Keyed join**: Selects the columns to perform the match based on keyed columns regardless of their names.To illustrate this method, we need to define keys in the same order for both tables.

```{r}
ProductsKeyed = setkey(copy(Products), id)
key(ProductsKeyed)

ProductReceivedKeyed = setkey(copy(ProductReceived), product_id)
key(ProductReceivedKeyed)

ProductsKeyed[ProductReceivedKeyed]
```

#### 3.1.3. Operations after joining

Most of the time after joining we need to make some additional transformations. To do so we have the following alternatives:

- Chaining a new instruction by adding a pair of brackets `[]`.
- Passing a list with the columns that we want to keep or create to the `j` argument.

Our recommendation is to use the second alternative if possible, as it is **faster** and uses **less memory** than the first one.

##### Managing shared column Names with the j argument

The `j` argument has great alternatives to manage joins with tables **sharing the same names for several columns**. By default all columns are taking their source from the the `x` table, but we can also use the `x.` prefix to make clear the source and use the prefix `i.` to use any column form the table declared in the `i` argument of the `x` table.

Going back to the little supermarket, after updating the `ProductReceived` table with the `Products` table, suppose we want to apply the following changes:

- Change the columns names from `id` to `product_id` and from `i.id` to `received_id`.
- Add the `total_value`.

```{r}
Products[
  ProductReceived,
  on = c("id" = "product_id"),
  j = .(product_id = x.id,
        name = x.name,
        price,
        received_id = i.id,
        date = i.date,
        count,
        total_value = price * count)
]
```


##### Summarizing with `on` in `data.table`

We can also use this alternative to return aggregated results based on the columns present in the `x` table.

For example, we might be interested in how much money we spend buying each product across days.

```{r}
dt1 = ProductReceived[
  Products,
  on = c("product_id" = "id"),
  by = .EACHI,
  j = .(total_value_received  = sum(price * count))
]

# alternative using multiple [] queries
dt2 = ProductReceived[
  Products,
  on = c("product_id" = "id"),
][, .(total_value_received  = sum(price * count)),
  by = "product_id"
]

identical(dt1, dt2)
```

#### 3.1.4. Joining based on several columns

So far we have just joined `data.table`s based on 1 column, but it's important to know that the package can join tables matching several columns.

To illustrate this, let's assume that we want to add the `tax_prop` from `NewTax` to **update** the `Products` table.

```{r}
NewTax[Products, on = c("unit", "type")]
```

### 3.2. Inner join

Use this method if you need to combine columns from 2 tables based on one or more references but ***keeping only rows matched in both tables***.

To perform this operation we just need to add `nomatch = NULL` to any of the prior join operations to return the same results.

```{r}
# First Table
Products[ProductReceived,
         on = c("id" = "product_id"),
         nomatch = NULL]

# Second Table
ProductReceived[Products,
                on = .(product_id = id),
                nomatch = NULL]
```

Despite both tables having the same information, there are some relevant differences:

- They present different column ordering.
- They have column name differences:
  - The `id` column in the first table has the same information as the `product_id` in the second table.
  - The `i.id` column in the first table has the same information as the `id` in the second table.

### 3.3. Anti-join

This method **keeps only the rows that don't match with any row of a second table**.

To apply this technique we can negate (`!`) the table located on the `i` argument.

```{r}
Products[!ProductReceived,
         on = c("id" = "product_id")]
```

As you can see, the result only has 'soda', as it was the only product that is not present in the `ProductReceived` table.

```{r}
ProductReceived[!Products,
                on = c("product_id" = "id")]
```

In this case, the operation returns the row with `product_id = 6,` as it is not present on the `Products` table.

### 3.4. Semi join

This method extracts **only the rows that match any row in a second table**, without combining the columns of the tables.

It's very similar to subset as join, but as in this time we are passing a complete table to the `i` we need to ensure that:

- Any row in the `x` table is duplicated due row duplication in the table passed to the `i` argument.

- All the renaming rows from `x` should keep the original row order. 


To make this, you can apply the following steps:

1. Perform a **inner join** with `which = TRUE` to save the row numbers related to each matching row of the `x` table.

```{r}
SubSetRows = Products[
  ProductReceived,
  on = .(id = product_id),
  nomatch = NULL,
  which = TRUE
]

SubSetRows
```

2. Select and sort the unique rows ids.

```{r}
SubSetRowsSorted = sort(unique(SubSetRows))

SubSetRowsSorted
```


3. Selecting the `x` rows to keep.

```{r}
Products[SubSetRowsSorted]
```
  

### 3.5. Left join

Use this method if you need to combine columns from 2 tables based on one or more references but ***keeping all rows present in the table located on the left***.

To perform this operation, we just need to **exchange the order between both tables** and the columns names in the `on` argument.

```{r}
ProductReceived[Products,
                on = list(product_id = id)]
```

Here some important considerations:

- **Column level**
  - The *first group* of columns now comes from the `ProductReceived` table as it is the `x` table.
  - The *second group* of columns now comes from the `Products` table as it is the `i` table.
  - It didn't add the prefix `i.` to any column.
  
- **Row level**
  - All rows from the `i` table were kept: the soda entry from `Products` that was not matched by any row in `ProductReceived` is still part of the results.
  - The row related to `product_id = 6` is not part of the results any more as it is not present in the `Products` table.


#### 3.5.1. Joining after chain operations

One of the key features of `data.table` is that we can apply several operations before saving our final results by chaining brackets.

```r
DT[
  ...
][
  ...
][
  ...
]
```

So far, if after applying all that operations **we want to join new columns without removing any row**, we would need to stop the chaining process, save a temporary table and later apply the join operation.

To avoid that situation, we can use special symbols `.SD`, to apply a **right join based on the changed table**.

```{r}
NewTax[Products,
       on = c("unit", "type")
][, ProductReceived[.SD,
                    on = list(product_id = id)],
  .SDcols = !c("unit", "type")]
```

### 3.6. Many to many join

Sometimes we want to join tables based on columns with **duplicated `id` values** to later perform some transformations later.

To illustrate this situation let's take as an example the `product_id == 1L`, which have 4 rows in our `ProductReceived` table.

```{r}
ProductReceived[product_id == 1L]
```

And 4 rows in our `ProductSales` table.

```{r}
ProductSales[product_id == 1L]
```

To perform this join we just need to filter `product_id == 1L` in the `i` table to limit the join just to that product and set the argument `allow.cartesian = TRUE` to allow combining each row from one table with every row from the other table.

```{r}
ProductReceived[ProductSales[list(1L),
                             on = "product_id",
                             nomatch = NULL],
                on = "product_id",
                allow.cartesian = TRUE]
```

Once we understand the result, we can apply the same process for **all products**.

```{r}
ProductReceived[ProductSales,
                on = "product_id",
                allow.cartesian = TRUE]
```

> `allow.cartesian` is defaulted to FALSE as this is seldom what the user wants, and such a cross join can lead to a very large number of rows in the result. For example, if Table A has 100 rows and Table B has 50 rows, their Cartesian product would result in 5000 rows (100 * 50). This can quickly become memory-intensive for large datasets.


#### 3.6.1. Selecting one match

After joining the table we might find out that we just need to return a single join to extract the information we need. In this case we have 2 alternatives:

- We can select the **first match**, represented in the next example by `id = 2`.

```{r}
ProductReceived[ProductSales[product_id == 1L],
                on = .(product_id),
                allow.cartesian = TRUE,
                mult = "first"]
```

- We can select the **last match**, represented in the next example by `id = 9`.

```{r}
ProductReceived[ProductSales[product_id == 1L],
                on = .(product_id),
                allow.cartesian = TRUE,
                mult = "last"]
```

#### 3.6.2. Cross join

If you want to get **all possible row combinations** regardless of any particular id column we can follow the next process:

1. Create a new column in both tables with a constant.

```{r}
ProductsTempId = copy(Products)[, temp_id := 1L]
```

2. Join both table based on the new column and remove it after ending the process, as it doesn't have reason to stay after joining.

```{r}
AllProductsMix =
  ProductsTempId[ProductsTempId,
                 on = "temp_id",
                 allow.cartesian = TRUE]

AllProductsMix[, temp_id := NULL]

# Removing type to make easier to see the result when printing the table
AllProductsMix[, !c("type", "i.type")]
```


### 3.7. Full join

Use this method if you need to combine columns from 2 tables based on one or more references ***without removing any row***.

As we saw in the previous section, any of the prior operations can keep the missing `product_id = 6` and the **soda** (`product_id = 4`) as part of the results.

To save this problem, we can use the `merge` function even though it is lower than using the native `data.table`'s joining syntax.

```{r}
merge(x = Products,
      y = ProductReceived,
      by.x = "id",
      by.y = "product_id",
      all = TRUE,
      sort = FALSE)
```


## 4. Non-equi join

A non-equi join is a type of join where the condition for matching rows is based on comparison operators other than equality, such as `<`, `>`, `<=`, or `>=`. This allows for **more flexible joining criteria**. In `data.table`, non-equi joins are particularly useful for operations like:

- Finding the nearest match.
- Comparing ranges of values between tables.

It is a great alternative when, after applying a right or inner join, you:

- Want to reduce the number of returned rows based on comparisons of numeric columns between tables.
- Do not need to retain the columns from table x *(the secondary `data.table`)* in the final result.

To illustrate how this works, let's focus on the sales and receives for product 2.
  
```{r}
ProductSalesProd2 = ProductSales[product_id == 2L]
ProductReceivedProd2 = ProductReceived[product_id == 2L]
```

If want to know, for example, you can find any receive that took place before a sales date, we can apply the following.

```{r}
ProductReceivedProd2[ProductSalesProd2,
                     on = "product_id",
                     allow.cartesian = TRUE
][date < i.date]
```

What does happen if we just apply the same logic on the list passed to `on`?

- As this operation is still a right join, it returns all rows from the `i` table, but only shows the values for `id` and `count` when the rules are met.

- The date related `ProductReceivedProd2` was omitted from this new table.

```{r}
ProductReceivedProd2[ProductSalesProd2,
                     on = list(product_id, date < date)]
```

Now, after applying the join, we can limit the results only showing the cases that meet all joining criteria.                                                               

```{r}
ProductReceivedProd2[ProductSalesProd2,
                     on = list(product_id, date < date),
                     nomatch = NULL]
```

### 4.1 Output column names in non-equi joins

When performing non-equi joins (`<`, `>`, `<=`, `>=`), column names are assigned as follows:

- The left operand (`x` column) determines the column name in the result.
- The right operand (`i` column) contributes values but does not retain its original name.
- By default, `data.table` does not retain the `i` column used in the join condition unless explicitly requested.

In non-equi joins, the left side of the operator (e.g., `x_int` in `x_int >= i_int`) must be a column from `x`, while the right side (e.g., `i_int`) must be a column from `i`.

Non-equi joins do not currently support arbitrary expressions (but see [#1639](https://github.com/Rdatatable/data.table/issues/1639)). For example, `on = .(x_int >= i_int)` is valid, but `on = .(x_int >= i_int + 1L)` is not. To perform such a non-equi join, first add the expression as a new column, e.g. `i[, i_int_plus_one := i_int + 1L]`, then do `.on(x_int >= i_int_plus_one)`.

```{r non_equi_join_example}
x <- data.table(x_int = 2:4, lower = letters[1:3])
i <- data.table(i_int = c(2L, 4L, 5L), UPPER = LETTERS[1:3])
x[i, on = .(x_int >= i_int)]
```

Key Takeaways:
- The name of the output column (`x_int`) comes from `x`, but the values come from `i_int` in `i`.
- The last row contains `NA` because no rows in `x` match the last row in `i` (`UPPER == "C"`).
- Multiple rows in `x` are returned to match the first row in `i` with `UPPER == "A"`.

If you want to keep the `i_int` column from `i`, you need to explicitly select it in the result:

```{r retain_i_column}
x[i, on = .(x_int >= i_int), .(i_int = i.i_int, x_int = x.x_int, lower, UPPER)]
```

Using prefixes (`x.` and `i.`) is not strictly necessary in this case since the names are unambiguous, but using them ensures the output clearly distinguishes `i_int` (from `i`) and `x_int` (from `x`).

If you want to exclude unmatched rows (an _inner join_), use `nomatch = NULL`:

```{r retain_i_column_inner_join}
x[i, on = .(x_int >= i_int), .(i_int = i.i_int, x_int = x.x_int, lower, UPPER), nomatch = NULL]
```

## 5. Rolling join

Rolling joins are particularly useful in time-series data analysis. They allow you to **match rows based on the nearest value** in a sorted column, typically a date or time column. 

This is valuable when you need to align data from different sources **that may not have exact matching timestamps**, or when you want to carry forward the most recent value. 

For example, in financial data, you might use a rolling join to assign the most recent stock price to each transaction, even if the price updates and transactions don't occur at the exact same times.


In our supermarket example, we can use a rolling join to match sales with the most recent product information.

Let's assume that the price for Bananas and Carrots changes at the first date of each month.

```{r}
ProductPriceHistory = data.table(
  product_id = rep(1:2, each = 3),
  date = rep(as.IDate(c("2024-01-01", "2024-02-01", "2024-03-01")), 2),
  price = c(0.59, 0.63, 0.65,  # Banana prices
            0.79, 0.89, 0.99)  # Carrot prices
)

ProductPriceHistory
```

Now, we can perform a right join giving a different price for each product based on the sale date.

```{r}
ProductPriceHistory[ProductSales,
                    on = .(product_id, date),
                    roll = TRUE,
                    j = .(product_id, date, count, price)]
```

If we just want to see the matching cases we just need to add the argument `nomatch = NULL` to perform an inner rolling join.

```{r}
ProductPriceHistory[ProductSales,
                    on = .(product_id, date),
                    roll = TRUE,
                    nomatch = NULL,
                    j = .(product_id, date, count, price)]
```

## 6. Taking advantage of joining speed

### 6.1. Subsets as joins

As we just saw in the prior section the `x` table gets filtered by the values available in the `i` table. This process is faster than passing a Boolean expression to the `i` argument.

To filter the `x` table at speed we don't need to pass a complete `data.table`, we can pass a `list()` of vectors with the values that we want to keep or omit from the original table.

For example, to filter dates where the market received 100 units of bananas (`product_id = 1`) or popcorn (`product_id = 3`) we can use the following:

```{r}
ProductReceived[list(c(1L, 3L), 100L),
                on = c("product_id", "count")]
```

As at the end, we are filtering based on a join operation the code returned a **row that was not present in original table**. To avoid that behavior, it is recommended to always add the argument `nomatch = NULL`.

```{r}
ProductReceived[list(c(1L, 3L), 100L),
                on = c("product_id", "count"),
                nomatch = NULL]
```


We can also use this technique to filter out any combination of values by prefixing them with `!` to negate the expression in the `i` argument and keeping the `nomatch` with its default value. For example, we can filter out the 2 rows we filtered before.

```{r}
ProductReceived[!list(c(1L, 3L), 100L),
                on = c("product_id", "count")]
```

If you just want to filter a value for a single **character column**, you can omit calling the `list()` function and pass the value to be filtered in the `i` argument.

```{r}
Products[c("banana","popcorn"),
         on = "name",
         nomatch = NULL]

Products[!"popcorn",
         on = "name"]
```

### 6.2. Updating by reference

Use `:=` to modify columns **by reference** (no copy) during joins. General syntax: `x[i, on=, (cols) := val]`.

**Simple One-to-One Update**  

Update `Products` with prices from `ProductPriceHistory`:

```{r}
Products[ProductPriceHistory, 
         on = .(id = product_id), 
         price := i.price]

Products
```

- `i.price` refers to price from `ProductPriceHistory`.
- Modifies `Products` in-place.

**Grouped Updates with `.EACHI`**

Get last price/date for each product:

```{r Updating_with_the_Latest_Record}
Products[ProductPriceHistory,
         on = .(id = product_id),
         `:=`(price = last(i.price), last_updated = last(i.date)),
         by = .EACHI]

Products
```

- `by = .EACHI` groups by i's rows (1 group per ProductPriceHistory row).
- `last()` returns last value

**Efficient Right Join Update**

Add product details to `ProductPriceHistory` without copying:

```{r}
cols <- setdiff(names(Products), "id")
ProductPriceHistory[, (cols) := 
  Products[.SD, on = .(id = product_id), .SD, .SDcols = cols]]
setnafill(ProductPriceHistory, fill=0, cols="price") # Handle missing values

ProductPriceHistory
```

- In `i`, `.SD` refers to `ProductPriceHistory`.
- In `j`, `.SD` refers to `Products`.
- `:=` and `setnafill()` both update `ProductPriceHistory` by reference.

## Reference

- *Understanding data.table Rolling Joins*: https://www.r-bloggers.com/2016/06/understanding-data-table-rolling-joins/

- *Semi-join with data.table*: https://stackoverflow.com/questions/18969420/perform-a-semi-join-with-data-table

- *Cross join with data.table*: https://stackoverflow.com/questions/10600060/how-to-do-cross-join-in-r

- *How does one do a full join using data.table?*: https://stackoverflow.com/questions/15170741/how-does-one-do-a-full-join-using-data-table

- *Enhanced data.frame*: https://rdatatable.gitlab.io/data.table/reference/data.table.html

